Decades of ``group polarization'' replications may be undermined by high false discovery rates

### 2024-11-18 GroupPolarizationStatmod/paper/ms.tex
good first draft of bar with markers for FWER, FDR

\usepackage{authblk}
\usepackage{etoolbox}
\makeatletter
\title{Foundational claims of group polarization are plausibly false,
with lessons for the empirical study of psychological dynamics}
\makeatother
\author[1,*]{{Matthew A.~Turner}}
\affil[1]{\small Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
\noindent \emph{Group polarization} is the name for a form of consensus
where members of a like-minded group become, on average, more extreme in
their opinions after discussing a topic. Group polarization is important
because it may drive political polarization if moderate partisans become
more extreme. Theoretically, group polarization is celebrated as a
regular, reliable outcome of a certain experimental design based on
dozens of published studies that purportedly demonstrate the phenomena.
However, in this paper we show that 55 of 60 published observations of
purported group polarization, from ten foundational studies, are equally
well-explained by a simple consensus model where group opinion variance
decreases, but extremity does not increase. We use a novel computational
meta-analysis framework designed to identify plausible pre- and
post-discussion opinion distributions that are consistent with simple
consensus, but appear to be group polarization when measured on ordinal,
Likert-style scales. This occurs due to ceiling effects induced by
ordinal measurements. At best this shows that a selection of
foundational research in social psychology must be discarded. At worst
this is the tip of the iceberg, and many similar social psychological
findings within and without the group polarization tradition will
similarly be shown to be unreliable.
\end{abstract}

@@

Since there seems to be no selection pressure on false discovery rates, 
experiment design for studying group polarization was free to evolve randomly.
Various choices for the number of bins and other instrumental details may have
then become entrenched in different communities defined by 
common theoretical perspectives taken by researchers in those communities. 
Therefore, when 
Isenberg (1986)~\cite{Isenberg1986} reports that his meta-analysis found more
support for the XX theory than the YY theory in the form of greater effect sizes,
it may just be random cultural drift that the XX community instruments were
more prone to greater .

>>>>

\usepackage{authblk}
\usepackage{etoolbox}
\makeatletter
\title{Decades of ``group polarization'' replications may be undermined by high false discovery rates}
% \title{Foundational claims of group polarization are plausibly false,
% with lessons for the empirical study of psychological dynamics}
\makeatother
\author[1,*]{{Matthew A.~Turner}}
\affil[1]{\small Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
\noindent 
\emph{Group polarization} is the name for a form of consensus where members of
a like-minded group become, on average, more extreme in their opinions after
discussing a topic. Group polarization is important because it may increase
social and political tensions if moderates become more extreme. Decades of
studies have replicated detections of group polarization, typically using an
ordinal scale to measure opinions.  However, these studies did not account for
ordinal measurements, which can result in spurious detections of opinion
change.  Lacking original data we can still calculate the probability that a
published detection of group polarization was spurious, i.e., to calculate the
\emph{false detection rate}.  Across 54 group polarization experimental
conditions from ten representative journal articles, we calculated false
detection rates to be between 0.52 and 0.88, with a median of 0.75, using a
generative model of group polarization experiments seeded with data from the
studies themselves.  We also use our model to develop experimental designs that
achieve an acceptable false discovery rate.  Much of group polarization
research may be unreliable, but this analysis can help change that. More
broadly, this work demonstrates one way that replication success alone does not
imply epistemic reliability.
\end{abstract}

@@

Since there seems to be no selection pressure on false discovery rates, 
experiment design for studying group polarization was free to evolve randomly.
Various choices for the number of bins and other instrumental details may have
then become entrenched in different communities defined by 
common theoretical perspectives taken by researchers in those communities. 
Therefore, when 
Isenberg (1986)~\cite{Isenberg1986} reports that his meta-analysis found more
support for the XX theory than the YY theory in the form of greater effect sizes,
it may just be random cultural drift that the XX community instruments had higher
intrinsic false detection rates. In some cases it may be debatable whether to
control for the Type I error rate, or may depend on context~\cite{Rubin2021}. 
In this context, however, it is clear that we must apply such selection pressure 
to group polarization experiments or it may be nearly impossible to trust 
statistical analyses of these studies.
